---
title: "Adding Data to an Existing Population Cube"
format: html
vignette: >
  %\VignetteIndexEntry{Adding Data to an Existing Population Cube}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
editor:
  markdown:
    wrap: 100
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

This vignette shows how to add newly released population data to an existing `poparray` cube. The
example uses the Texas Demographic Center county population estimates workflow from
`data-raw/tdc_estimates.r`.

The common use case is an annual update. For example, an existing cube may already contain TDC
estimate years through 2023, and a new file named something like
`2024_ASRE_Estimate_alldata.csv` has been downloaded. Rather than rebuilding the whole cube from
all historical source files, `add_population_data()` can read only the new file, validate it against
the existing cube, and write an updated cube with 2024 appended to the `year` dimension.

The function does not edit the HDF5 dataset in place. It creates an expanded cube, copies the
existing cube by blocks, writes the new slice, writes updated metadata, and then replaces the source
file when requested. This keeps the existing HDF5 data lazy during the update and avoids realizing
the full old cube in memory.

# What the User Must Provide

To add data, the user needs four things:

1. An existing population cube, either as a file path or as a `series_id` known to
   `open_poparray()`.
2. A reader function that returns the newly downloaded source table.
3. A transformer function that converts the source table to the same canonical dimensions as the
   existing cube.
4. The dimension order used by the cube.

For the TDC estimate cube, the dimensions are:

```{r dimensions}
dims <- c("year", "area.name", "sex", "age.char", "race.eth")
```

The new file must produce records with those five dimension columns plus the value column,
`population`.

# Setup

In a development workflow, the package and source-data paths are usually initialized by
`data-raw/control_def.r`. An interactive session typically looks like this:

```{r setup}
library(tarr.pop)
library(data.table)
library(tidyverse)
library(janitor)
library(rage)

devtools::load_all()

cube_root <- tarr.pop::init_cubes()
tdc_estimates_file <- file.path(cube_root, "base", "tdc_estimates_county.h5")
```

The source CSV folder should point to the folder where the TDC estimate files were downloaded:

```{r source-dir}
input_dir <- file.path(
  tarr::paths$population,
  "Estimates",
  "Texas Demographic Center",
  "asre"
)
```

# Reader Function

The reader should return only the new source data. For an annual update, this usually means changing
the file pattern to match the newly downloaded file.

The `read_tdc_estimates_raw()` function from `data-raw/tdc_estimates.r` already supports a
`pattern` argument, so the same reader can be reused:

```{r read-2024}
raw_2024 <- read_tdc_estimates_raw(
  pattern = "2024_ASRE_Estimate_alldata\\.csv",
  input_dir = input_dir
)
```

When using `add_population_data()`, do not call the reader yourself. Pass the reader function and
the file-matching arguments to the API:

```{r reader-api}
reader <- read_tdc_estimates_raw
```

# Transformer Function

The transformer must return the same dimensions as the existing cube. For the TDC estimate cube,
`transform_tdc_estimates()` performs the source-specific work:

```{r transformer}
transformer <- transform_tdc_estimates
```

The transformer is responsible for:

- converting the wide TDC race/sex columns to long format
- creating `race.eth` and `sex`
- normalizing county names into `area.name`
- normalizing age labels into `age.char`
- removing the Texas total when the cube is county-only
- dropping records where `population` is `NA`
- removing physical `"All"` aggregate rows

The last point is important. In `tarr.pop`, `"All"` is treated as a virtual aggregate. It should not
be physically stored as a dimension label in the cube.

# Add the New Year

The update call is:

```{r add-2024}
add_population_data(
  cube = tdc_estimates_file,
  reader = read_tdc_estimates_raw,
  transformer = transform_tdc_estimates,
  dims = dims,
  add_dim = "year",
  pattern = "2024_ASRE_Estimate_alldata\\.csv",
  input_dir = input_dir,
  counties = default_counties,
  include_texas_total = FALSE,
  source_meta = list(
    note = "Texas Demographic Center county estimates, updated with 2024",
    source = "Texas Demographic Center, Estimates program",
    population_type = "Estimate"
  )
)
```

This call reads and transforms only the 2024 file. The existing cube supplies the valid labels for
`area.name`, `sex`, `age.char`, and `race.eth`. The new data supply the new `year` label.

By default, `add_population_data()` replaces the input file when `cube` is a file path. Replacement
is done safely through a temporary HDF5 file in the same folder. If the write succeeds, the old file
is removed and the temporary file is renamed into place.

# Write to a Different File

During testing, it is often safer to write the updated cube to a new file:

```{r add-2024-new-file}
updated_file <- file.path(cube_root, "base", "tdc_estimates_county_with_2024.h5")

add_population_data(
  cube = tdc_estimates_file,
  reader = read_tdc_estimates_raw,
  transformer = transform_tdc_estimates,
  dims = dims,
  add_dim = "year",
  output_filepath = updated_file,
  overwrite = TRUE,
  pattern = "2024_ASRE_Estimate_alldata\\.csv",
  input_dir = input_dir,
  counties = default_counties,
  include_texas_total = FALSE,
  source_meta = list(
    note = "Texas Demographic Center county estimates, test update with 2024",
    source = "Texas Demographic Center, Estimates program",
    population_type = "Estimate"
  )
)
```

Use this form when testing a new source file or transformer change. Once the result is inspected,
the same call can be run without `output_filepath` to replace the production cube.

# Completion Policy

The `completion_policy` argument controls how missing cells in the new slice are handled.

With `completion_policy = "error"`, every valid cell must be present in the new data. If a county,
sex, age, race, and year combination is missing, the function stops and reports examples of missing
records.

```{r completion-error}
add_population_data(
  cube = tdc_estimates_file,
  reader = read_tdc_estimates_raw,
  transformer = transform_tdc_estimates,
  dims = dims,
  add_dim = "year",
  completion_policy = "error",
  pattern = "2024_ASRE_Estimate_alldata\\.csv",
  input_dir = input_dir
)
```

With `completion_policy = "na"`, missing valid cells are written as `NA`. This is useful when the
source program has a known structural gap. For example, older TDC files did not include every
`race.eth` category in every year.

```{r completion-na}
add_population_data(
  cube = tdc_estimates_file,
  reader = read_tdc_estimates_raw,
  transformer = transform_tdc_estimates,
  dims = dims,
  add_dim = "year",
  completion_policy = "na",
  pattern = "2024_ASRE_Estimate_alldata\\.csv",
  input_dir = input_dir
)
```

With `completion_policy = "zero"`, missing valid cells are filled with zero. This should be used
only when zero is a defensible structural value, not merely because a source record is absent.

# Support Table

When `support = NULL`, `add_population_data()` derives a support table from the existing cube. For a
new year, that means crossing the new year label with the existing labels for all other dimensions.
This is usually the correct behavior for annual updates, because the existing cube defines the
valid cell space.

If the new source file intentionally changes the valid cell space, provide an explicit support
table:

```{r explicit-support}
support_2024 <- tidyr::expand_grid(
  year = 2024,
  area.name = areas(open_poparray("tdc_estimates_county")),
  sex = sexes(open_poparray("tdc_estimates_county")),
  age.char = ages(open_poparray("tdc_estimates_county")),
  race.eth = c("white", "black", "asian", "other", "hispanic")
)

add_population_data(
  cube = tdc_estimates_file,
  reader = read_tdc_estimates_raw,
  transformer = transform_tdc_estimates,
  dims = dims,
  add_dim = "year",
  completion_policy = "na",
  support = support_2024,
  pattern = "2024_ASRE_Estimate_alldata\\.csv",
  input_dir = input_dir
)
```

This example uses `open_poparray()` to retrieve existing labels. That is metadata-oriented and does
not require realizing the full cube.

# Safeguards

`add_population_data()` intentionally refuses some updates.

It stops if the new data contain an existing `year` label. This prevents accidentally overwriting
or duplicating a year already in the cube.

```{r overlap-example}
add_population_data(
  cube = tdc_estimates_file,
  reader = read_tdc_estimates_raw,
  transformer = transform_tdc_estimates,
  dims = dims,
  add_dim = "year",
  pattern = "2023_ASRE_Estimate_alldata\\.csv",
  input_dir = input_dir
)
```

If 2023 is already in the cube, this call errors.

It also stops if the new data contain labels in non-appended dimensions that are not already in the
cube. For example, if a new `race.eth` label appears in the 2024 source file, the cube's dimension
contract has changed. In that case, decide whether the cube should be rebuilt with a new schema or
whether a source-specific transformer should map the new label into the existing label set.

# Inspect the Result

After adding the new data, open the cube and inspect its metadata:

```{r inspect}
updated <- open_poparray("tdc_estimates_county")

years(updated)
areas(updated)[1:5]
sexes(updated)
ages(updated)[1:10]
```

Small slices can be converted to a data frame for checking. This is an **EAGER** operation because
`as.data.frame()` realizes the selected slice. Filter or subset first.

```{r inspect-slice}
check_2024 <- updated["2024", , , , , drop = FALSE]

as.data.frame(check_2024) |>
  dplyr::filter(area.name == "Tarrant") |>
  dplyr::slice_head(n = 10)
```

For large cubes, avoid converting the full object to a data frame. Work with selected years,
selected areas, or other small slices.

# Relationship to Full Ingestion

Use `ingest_population()` when creating a cube from source files.

Use `add_population_data()` when the cube already exists and the source update adds new labels along
one dimension, usually `year`.

If a new source file changes multiple dimensions at once, such as adding a new race category and a
new year, a full rebuild may be the cleaner workflow. The cube schema is part of the data contract,
and changing that schema should be deliberate.

