---
title: "API Reference tarr.pop "
format: html
vignette: >
  %\VignetteIndexEntry{tarr.pop API Reference}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor: 
  markdown: 
    wrap: 100
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(dplyr)
library(tarr.pop)
```

This document is the long-form reference for the user-facing API. It is organized around the way
population cubes are created, inspected, transformed, summarized, projected, and stored. Each
section also explains how the functions are intended to be used together. 

The package is built around delayed (lazy), dimension-aware population arrays - `poparrays`. As most of the arrays have
three or more dimensions they are often referred to as 'cubes'. In normal use, a `poparray` should remain lazy for as
long as possible so that filtering, grouping, and many summaries can be expressed without materializing the full cube in
memory. When the cube is materialized, only the needed data is retrieved from the cube. This matters both for
performance and for correctness: The package is designed to preserve dimension names, roles, and semantic metadata while
preventing unsafe aggregations across dimensions with overlapping categories. For example, census bureau race categories
can be overlapping with a single "race" and "race in combination" like "Asian" and "Asian or in combination" where you would
not not want to naively `sum()` across as totals would include duplicates.

Two ideas appear throughout this reference: 

1. A `poparray` is not just an array of numbers. It is an array plus dimensional meaning, including time, area, and category semantics. Every poparray must have at least two dimensions that encode the time and area roles. 


2. Long-table or tibble conversions are often convenient, but they are intentionally marked **EAGER** because they
realize cube values into a tabular structure.

## Arrays, Data Frames, Lazy Work, and Eager Work 

Many R users spend most of their time with data frames, so it is worth being explicit about how an array-based workflow
differs. A data frame is organized as rows and columns. Each row is usually treated as one observation, and each column
is treated as one variable. That is an excellent general-purpose structure, but it is not always the most natural
structure for population data. 

Population data is often defined by several dimensions at once: year, area, age, sex, race, and ethnicity are all part
of what a value means. In a data frame, that multi-dimensional structure is present, but it is implicit. In an array,
the structure is explicit. 

One way to picture the difference is this:

``` text
Data frame structure

year   area      age      sex     population
2020   Tarrant   0-4      F       73421
2020   Tarrant   0-4      M       76810
2020   Tarrant   5-9      F       75103
2020   Dallas    0-4      F       181552
...

Each row stores one combination of categories. Notice how the columns year, 
area, age, and sex repeat the same information. The dimensional 
structure is carried by repeated values in the columns.
```

``` text
Array structure

value[year, area, age, sex]

                 sex
              F      M
year area age
2020 Tarr 0-4 73421 76810
          5-9 75103 78744
     Dall 0-4 181552 190331

Each axis is a named dimension - year, area, age and sex.  The same as the data frame 
columns, however those labels are not repaeted, rather named only once. Each cell is 
addressed by its position on every dimension at once.
```

The most important conceptual shift from data frames to arrays is this: arrays do not start from rows.
They start from dimensions. That means the structure of the data is not reconstructed from repeated
columns each time you work with it. The structure is already built into the object. Another 
important difference betwewn the two is that arrays are composed one of one data type - logical, integer, double, or character whereas 

This matters for both correctness and efficiency. In a data frame, each row repeats the labels
needed to identify the cell: the year value appears again and again, the area name appears again and
again, etcera. In an array, those labels are stored once per dimension and the values live in the multi-dimensional grid. That is much more compact, especially when the same categories repeat across many combinations.

Another way to picture the memory difference is:

``` text
Data frame

row 1: "2020", "Tarrant", "0-4", "F", 73421
row 2: "2020", "Tarrant", "0-4", "M", 76810
row 3: "2020", "Tarrant", "5-9", "F", 75103
...

The identifying labels are repeated in every row.
```

``` text
Array

dimnames$year = c("2020", "2021", ...)
dimnames$area = c("Tarrant", "Dallas", ...)
dimnames$age  = c("0-4", "5-9", ...)
dimnames$sex  = c("F", "M")

values = one numeric block arranged on those dimensions

The identifying labels are stored once per axis.
```

The package builds on that compact array representation and then goes a step further by supporting
delayed computation. This is where the terms **lazy** and **EAGER** become important.

In this package, **lazy** means that an operation is described, recorded, or planned without
immediately realizing all values into ordinary in-memory R objects. A lazy filter, grouped
operation, or summary can often stay as a DelayedArray-style computation, which allows the package
to work efficiently with very large disk-based cubes.

By contrast, **EAGER** means that the operation materializes data now. When a `poparray` is
converted to a data frame or tibble, the package expands the cube into rows and attaches the
identifying labels to each row. That is often exactly what the user wants for export, printing,
modeling, or plotting, but it is more expensive in both memory and computation.

The contrast can be summarized like this:

``` text
Lazy
  Keep the cube in array form
  Keep labels on dimensions
  Delay reading or computing full results when possible
  Best for filtering, regrouping, summaries, and storage-backed workflows

EAGER
  Materialize values now
  Often expand the cube into long rows
  Useful for reporting, export, and table-oriented tools
  More memory-intensive for large cubes
```

For an R user used to data frames, a good rule is: stay in array form while you are doing
population-cube work, and move to a data frame only when you need a data-frame result. That is why
`tarr.pop` treats long-table conversion as important but not primary. The array is the working
representation; the data frame is usually the delivery representation.

## Working Principles

Most workflows follow a common pattern:

1.  Open or construct a cube.
2.  Inspect its roles, dimensions, categories, and semantics.
3.  Filter or regroup dimensions without realizing values.
4.  Compute summaries or create projections.
5.  Convert to a data frame or tibble only when tabular output is actually needed.

The package design tries to prevent common mistakes. For example, age is treated as interval data rather than plain
strings, and grouping/collapse operations are expected to respect semantic safety rather than summing across overlapping
labels by accident. 

```{r open_inspect}

# Open a poparray
census_est <- open_tarr_pop(population$census.bureau$estimates)

census_est

# dimensions and names
dim(census_est)
dimnames(census_est) |> names()


dimnames(census_est)["race"]

```

## Core Cube Objects

The starting point for most work is the `poparray` class. A `poparray` represents a population cube
together with the metadata needed to interpret its dimensions safely. The projection workflow
introduces a second class, `poparray_projection`, which keeps projected values and related
uncertainty in a structured form.

| Function | Purpose | Object Class |
|---------------------------------|---------------------------------|---------------------------------|
| `new_poparray()` | Construct a `poparray` from an HDF5-backed delayed array. | `poparray` |
| `as.poparray()` | Coerce supported objects to `poparray`. | `poparray` |
| `is.poparray()` | Test whether an object is a `poparray`. | logical / `poparray` |
| `poparray_projection()` | Construct a `poparray_projection`. | `poparray_projection` |

Use `new_poparray()` when you are assembling a cube with metadata explicitly and want a canonical
object. Use `as.poparray()` when you already have a compatible object and want to normalize it to
package conventions. In either case, the important point is that construction should preserve
delayed storage, dimension names, roles, and semantic metadata.

## Cube Accessors

Accessors expose the metadata that makes a `poparray` more than a numeric array. These functions are
usually the first thing to inspect when you are trying to understand a cube, verify assumptions in a
script, or document how downstream code interprets each dimension.

| Function | Purpose | Object Class |
|---------------------------------|---------------------------------|---------------------------------|
| `time_role()` | Get the time dimension name. | `poparray` |
| `area_role()` | Get the area dimension name. | `poparray` |
| `data_col()` | Get the value-column name used for tabular coercion. | `poparray`, `poparray_projection` |
| `data_col<-()` | Set the value-column name. | `poparray`, array-like |
| `dim_semantics()` | Get per-dimension semantic metadata. | `poparray` |
| `get_source()` | Get source/provenance metadata. | `poparray` |

These accessors matter because many higher-level operations rely on named dimensions rather than
positional indexing. A workflow that depends on "dimension 1 means year" is fragile; a workflow that
asks for the time role directly is much safer and easier to read. The same principle applies to
source metadata and the value-column name when converting to a data frame or tibble.

```{r accessors}
# get the dimnesion names that are in the time and area roles
time_role(census_est)
area_role(census_est)

# when converted to a data frame, what will the data column be cllaed?
data_col(census_est)

# data source
get_source(census_est)

```


## Dimension Label/Cateogry Accessors

Label accessors provide a readable way to inspect valid values along standard dimensions such as
year, area, age, sex, race, and ethnicity. They are plural forms of the actual dimension  name. 
The accessors are only valid if the cube actually has a dimension name that matches. 
They are useful both in interactive analysis and in code that needs to validate user choices before subsetting or regrouping.  

| Function        | Purpose                              | Object Class           |
|-----------------|--------------------------------------|------------------------|
| `years()`       | Return labels for `year`.            | `poparray`, array-like |
| `areas()`       | Return labels for `area.name`.       | `poparray`, array-like |
| `ages()`        | Return labels for `age.char`.        | `poparray`, array-like |
| `sexes()`       | Return labels for `sex`.             | `poparray`, array-like |
| `races()`       | Return labels for `race`.            | `poparray`, array-like |
| `ethnicities()` | Return labels for `ethnicity`.       | `poparray`, array-like |
| `dim_labels()`  | Return labels for a named dimension. | `poparray`, array-like |

`dim_labels()` is the general interface, while helpers such as `years()` and `ages()` are optimized
for common dimensions in demographic workflows. Age labels should be interpreted as intervals, not
as arbitrary strings. That distinction is important when filtering ranges or creating grouped age
bands.

```{r label_accessors}
# dim_labels works for any dimension that exists in a cubes 
dim_labels(arr = census_est, "year")

# some accessors are pre-defined for some dimensions
areas(census_est) |> head()

years(census_est)
```


## Filtering and Subsetting

Filtering narrows a cube by label while preserving the dimensional structure that remains. In normal
workflows, this should stay lazy: the goal is to describe the slice you want, not to immediately
materialize it.

| Function | Purpose | Object Class |
|---------------------------------|---------------------------------|---------------------------------|
| `filter.poparray()` | Restrict dimensions by label predicates without materializing data. | `poparray` |
| `%between%` | Range helper for ordered dimension filtering. | predicate helper |
| `split.poparray()` | Split a cube into slices by one dimension. | `poparray` |
| `by.poparray()` | Apply a function by dimension groups. | `poparray` |

This family of functions is useful when you want to focus on a subset of years, a region, or an age
interval before computing summaries or plots. The main safety benefit is that subsetting remains
dimension-aware. Using label predicates is more robust than numeric positions, especially when the
cube has been filtered or regrouped earlier in the workflow.

`%between%` is especially helpful for ordered dimensions such as time or age intervals.
`split.poparray()` and `by.poparray()` support more structured iteration without discarding the
semantics that individual slices still need.

```{r filter}
# reduce the numbe rof counties and years
tarrant_est <- census_est |> 
  filter(area.name  %in% "Tarrant",
         year %between% c(2019, 2022))

tarrant_est

areas(tarrant_est)

years(tarrant_est)

```


## Grouped Aggregation

Aggregation is where semantic errors become easy to make, so the grouped aggregation API is
deliberately explicit. These functions are meant to collapse a dimension category/labels into groups
while checking that the operation is meaningful.

| Function         | Purpose                                                                 | Object Class |
|---------------------------------|---------------------------------|---------------------------------|
| `collapse_dim()` | Collapse one dimension into group with semantic safety checks. |
| `group_ages()`   | Age-specific grouped collapse built on `collapse_dim()`.       |

`collapse_dim()` is the general tool for regrouping labels. `group_ages()` is the specialized helper
for age intervals, where overlap and ordering matter especially strongly. These functions are
central to the package's design because a plain array sum can produce numerically valid but
substantively wrong results. The grouped API exists to reduce that risk.

Users should prefer these helpers to ad hoc recoding when building broad age bands or collapsing
categories for analysis. The aim is not only to simplify code, but also to preserve metadata and
guard against unsafe aggregation.

## Summaries and Coercion

This section combines two related ideas: computing summaries from a cube and coercing the cube into
other representations. The distinction to keep in mind is that many summaries can remain delayed or
block-wise, while tabular coercion is usually **EAGER**.

| Function | Purpose | Object Class |
|---------------------------------|---------------------------------|---------------------------------|
| `summary.poparray()` | Return summary statistics for a `poparray`. | `poparray` |
| `as.data.frame.poparray()` | Convert a `poparray` to a long data frame. **EAGER**. | `poparray` |
| `as_tibble.poparray()` | Convert a `poparray` to a tibble. **EAGER**. | `poparray` |
| `as.double.poparray()` | Convert cube values to double while preserving cube structure. | `poparray` |
| `sd()` | Standard deviation for `poparray` using delayed block reduction. | `poparray` |
| `sum()` | Semantically guarded sum for `poparray`. | `poparray` |

`summary.poparray()`, `sd()`, and `sum()` are designed for analytical work on the cube itself. In
particular, `sum()` is semantically guarded, which reflects the package's broader goal of preventing
invalid aggregation across overlapping dimensions or categories.

`as.data.frame.poparray()` and `as_tibble.poparray()` are included because many downstream tools
expect long-form tables. They are intentionally useful, but they should be treated as endpoints
rather than default intermediate steps. Once you convert a large delayed cube to a table, you lose
the memory and performance advantages of lazy array operations. Those conversions are best reserved
for reporting, export, or compatibility with tools that require tabular input.

For users who want a tidyverse workflow, `as_tibble.poparray()` provides that entry point, but it
should still be understood as **EAGER**. Tidyverse code is often row-wise and table-oriented,
whereas `tarr.pop` is designed around dimension-aware arrays. The conversion is valuable, but it
changes the computational model.

## Projection API

The projection API supports forecasting workflows by separating observed cube data from projected
values and associated uncertainty. This allows projections to be stored, inspected, converted, and
plotted through a consistent interface.

| Function | Purpose | Object Class |
|---------------------------------|---------------------------------|---------------------------------|
| `project()` | Build a projection from a `poparray`. | `poparray` -\> `poparray_projection` |
| `projection()` | Extract projected values. | `poparray_projection` |
| `std_error()` | Extract standard errors. | `poparray_projection` |
| `as.poparray.poparray_projection()` | Coerce a projection to `poparray`. | `poparray_projection` -\> `poparray` |
| `as.data.frame.poparray_projection()` | Convert projection to tabular output. **EAGER**. | `poparray_projection` |
| `as_tibble.poparray_projection()` | Convert projection to tibble output. **EAGER**. | `poparray_projection` |
| `plot.poparray_projection()` | Plot projections as time series or pyramids. | `poparray_projection` |

`project()` is the entry point for creating a `poparray_projection` from an existing population
cube. Once created, `projection()` and `std_error()` make the structure explicit by exposing the
main projected values and uncertainty components separately.

As with base cubes, tabular coercions for projections are **EAGER** and are best used when a
long-form output is required. `plot.poparray_projection()` provides a higher-level way to inspect
projected trajectories without forcing the user to manually reshape the object first.

## Cube Storage and Persistence

Storage functions manage where cubes live on disk and how they are created, saved, and reopened in
the canonical HDF5-backed format. This layer is what allows package workflows to stay lazy across
sessions instead of being limited to in-memory arrays.

| Function                  | Purpose                                                | Object Class      |
|---------------------------|--------------------------------------------------------|-------------------|
| `cube_path()`             | Get the active cube storage root.                      | storage config    |
| `set_cube_path()`         | Set the active cube storage root.                      | storage config    |
| `init_cubes()`            | Initialize cube storage directories and bundled cubes. | storage config    |
| `build_cube_if_missing()` | Create a cube only if not already present.             | storage / builder |
| `create_poparray()`       | Create and save a cube in canonical HDF5 schema.       | `poparray`        |
| `save_poparray()`         | Save a `poparray` to canonical HDF5 schema.            | `poparray`        |
| `open_poparray()`         | Open a registered cube lazily.                         | `poparray`        |
| `open_tarr_pop()`         | Backward-compatible alias for `open_poparray()`.       | `poparray`        |

In a typical workflow, `init_cubes()` prepares the storage location, `create_poparray()` or
`save_poparray()` writes data using the canonical schema, and `open_poparray()` reopens it lazily
for analysis. The distinction between creation, saving, and opening matters because the package
wants the on-disk representation to remain consistent with the metadata contracts used by
`poparray`.

This storage layer also reduces a common mistake in array-heavy work: losing track of which object
is authoritative. By using canonical paths and open/save helpers, workflows can remain reproducible
and easier to document.

## Metadata Administration

Most users will read metadata more often than they write it. Metadata write operations are
deliberately treated as elevated administration because they affect how a cube is interpreted
throughout the package.

These functions are intended for elevated metadata administration. Write/edit operations require
`options(tarr.pop.metadata_role = "admin")` or `Sys.setenv(TARR_POP_METADATA_ROLE = "admin")`.

| Function | Purpose | Object Class |
|---------------------------------|---------------------------------|---------------------------------|
| `dim_semantics()` | Read or write `dim_semantics` metadata for a `poparray` or cube file. | poparray or cube metadata |
| `roles()` | Read or write role metadata for a `poparray` or cube file. | poparray or cube metadata |
| `source_meta()` | Read or write source/provenance metadata for a `poparray` or cube file. | poparray or cube metadata |
| `data_col()` | Read or write the value-column name for a `poparray` or cube file. | poparray or cube metadata |
| `cube_metadata()` | Read or write the bundled canonical metadata in one validated operation. | poparray or cube metadata bundle |

The administrative boundary is intentional. Changing roles, semantics, or source metadata can
silently alter the meaning of later analyses if done casually. Requiring an explicit admin mode
helps reduce accidental edits and signals that these functions affect package-level invariants, not
just local object convenience.

When possible, prefer bundled or validated metadata operations over piecemeal manual editing. That
keeps related fields synchronized and makes mistakes easier to catch.

## Data Conversion Helpers

The helper conversion functions bridge array and data-frame representations. They are useful for
import, export, and interoperability, but they should be used with awareness of the computational
model being crossed.

| Function       | Purpose                           | Object Class         |
|----------------|-----------------------------------|----------------------|
| `df_2_array()` | Convert a data frame to an array. | data frame -\> array |
| `array_2_df()` | Convert an array to a data frame. | array -\> data frame |

These helpers are lower-level than `as.data.frame.poparray()` or `as_tibble.poparray()`. They are
especially useful when constructing examples, tests, or import pipelines that need to move between
tabular source data and array form before wrapping that array in a `poparray`.

Because they sit close to the representation boundary, they are also a place where users can make
mistakes such as dropping labels, confusing dimension order, or flattening interval-valued age
categories into plain text. When these helpers are used in package code or examples, the resulting
arrays should be checked carefully before being treated as semantically complete population cubes.

## Choosing Between Lazy and Eager Workflows

For most package operations, the recommended approach is:

-   Keep data as a `poparray` while filtering, regrouping, summarizing, and projecting.
-   Use named dimensions and metadata accessors rather than positional logic.
-   Convert to a data frame or tibble only when a table is genuinely needed for output, reporting,
    or compatibility.

Base R alternatives are sometimes useful, especially for simple reporting or inspection, but they
should be labeled **EAGER** when they force realization. The same principle applies to tidyverse
workflows: they can be practical, but they usually assume row-wise tabular data and therefore move
the workflow out of the package's lazy array model.

## Common Mistakes This API Tries to Prevent

Several aspects of the API are intentionally opinionated because demographic array work is easy to
get wrong in subtle ways:

-   Summing across overlapping groups can produce invalid totals.
-   Treating age labels as ordinary strings can break ordered grouping logic.
-   Relying on positional dimensions makes code brittle after filtering or regrouping.
-   Converting to long tables too early can cause unnecessary realization and memory growth.
-   Editing cube metadata casually can change analytical meaning without changing numeric values.

The API design addresses these risks by emphasizing semantic metadata, dimension roles, guarded
aggregation, lazy storage, and explicit administrative boundaries for metadata edits.
