---
title: "Introduction to tarr.pop"
format: html
execute:
  cache: true
vignette: >
  %\VignetteIndexEntry{Introduction to tarr.pop}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| message: false
#| warning: false
#| echo: false
 
library(tarr.pop)
library(dplyr)
library(tibble)
library(ggplot2)
```

# Introduction

The `tarr.pop` package is designed for working with population values as
multi-dimensional data rather than long tables of rows. The distinction
is important in epidemiology and demography work flows, where a
population count is usually defined by several dimensions at once, for
example:

-   year
-   geography
-   age
-   sex
-   race
-   ethnicity

In a regular table, those dimensions are present, but their structure is
only implied by columns. In `tarr.pop`, the same data is treated as a
population cube, which makes the structure explicit and easier to work
with safely. In R parlance a cube is a multi-dimensional
[array](https://www.programiz.com/r/arrays).

Each dimension can have a number of labeled categories/levels. For
example, a `sex` dimension would have "female" and "male". The area
dimension could have an entry for each county or other divisions like
census block. The year dimension would have entries for each year. In
the case of the decennial census, those years could be 2000, 2020, and
2020.

# Why use a population cube?

Population data is naturally multi-dimensional. A single cell in a cube
might mean:

> the number of people in Tarrant County, in 2022, age 65 to 69, female

That is not a row-first idea. It is a combination of labeled dimensions.

# The `poparray` class

Central to how the `tarr.pop` package represents population data is the
`poparray` class to handle the dimensional structure. `poparray` builds
on two other classes from the Bioconductor ecosystem DelayedArray for
delaying array actions for optimization, and HDF5Array to hanlde
operations on disk rather than memory. This results in large cubes of
data stay on disk and are manipulated "lazily" (deferred actions) until
you actually need a realized (retrieved) result.

A `poparray` is an R S4 class that stores:

-   the population values themselves
-   the names and labels of each dimension
-   the role of key dimensions such as time and area
-   semantic information about each dimension

This means a `poparray` is more than an array of numbers. It is an array
plus a contract describing what those numbers mean by dimension.

## Roles

Every `poparray` has:

-   one time dimension
-   one area dimension
-   zero or more additional stratification dimensions

Those roles are stored with the object, so functions can work with the
cube by meaning rather than by hard-coded position. The additional
stratification dimensions mentioned above may include things like age,
race, ethnicity, house size, income, education, etc.

Several cubes are included with the package. Sources include the US
Census Bureau, National Cancer Institute SEER program, and the Texas
Demographic Center. To work with a cube of estimates from the Texas
Demographic Center use the `open_array()` function like so:

```{r open_tdc_estimates}
# open the Texas Demographic Center estimates cube.
pop <- open_poparray("tdc_estimates_county")

# Look at the pop cube information 
pop
```

The printed information for `pop` includes the series name, the data
source, last cube update, length is the number of values, the time and
area role names, dimension names with number of levels, and column name
used when converted to a data frame.

## `DimSemantics`

The `DimSemantics` class stores semantic information for each dimension
in a cube. The `poparray` class uses objects of `DimSemantics` to enable
operations like collapsing levles n a dimension or preventing sum() when
overlapping levles exist. Dimdemtics describe whether a dimension
behaves like:

-   a partition, where categories are mutually exclusive
-   a set, where categories may overlap
-   an interval scale, such as age groups or years

That matters because not every aggregation is safe. Some categories
overlap and should not be summed without care. By storing semantics with
the cube, `poparray` can help prevent invalid operations.

You can inspect the semantic metadata directly of the pop variable
created above :

```{r explore_semantics}

names(dim_semantics(pop))  # names of the dimensions

dim_semantics(pop)$year    

dim_semantics(pop)$age.char
```

# A first workflow

`poparray` is meant handle a large data cube on the backend and allow
you to specify the data you need. Here is an example workflow with some
steps along the way so you can see what is happening:

1.  open a cube
2.  inspect the structure
3.  filter to a smaller analysis slice
4.  collapse remaining stratification dimensions
5.  convert to a tibble for plotting

## Open a cube

For this example, use the Texas Demographic Center county estimates
cube.

```{r fkow_cube_open}
pop <- open_poparray("tdc_estimates_county")
pop
```

Notice the number of data elements in the cube and the size of each
dimension. Let's see some of the cube meta data. Start by asking a few
basic questions about the cube.

```{r meta_data}
dim(pop)              # size of each dimension.
names(dimnames(pop))  # names of each dimension.
time_role(pop)        # name used for the time diomension.
area_role(pop)        # name used for the area dimension.
```

The dimension labels are also easy to inspect and contain information in
differenmt "slots".

```{r}
head(dimnames(pop)[[time_role(pop)]])

head(dimnames(pop)[[area_role(pop)]])
```

## Filter to a smaller slice

Suppose we want to compare Dallas, Tarrant, and Travis counties over the
most recent five years in the cube.

```{r small_explore}
# Names of the counties
target_counties <- c("Dallas", "Tarrant", "Travis")

recent_years <- dimnames(pop)[[time_role(pop)]] |>
  tail(5)

pop_small <- pop |>
  dplyr::filter(
    area.name %in% target_counties,
    year %in% recent_years,
    age.char %in%  ages(pop, remove = "85 +")
  )

pop_small
```

This filtering step is still lazy, meaning the actions are queued to
occur. At this point, `poparray` has created a smaller cube view, but it
has not converted the full result into an in-memory data frame. You can
still look at what to expect to see in the smaller cube view. Compare
the number of values in pop_small
`r length(pop_small) |> format(big.mark = ",")` to that in pop
`r length(pop) |> format(big.mark = ",")`. Just by reducing the number
of `levels` in area.name and `year` the number of values are smaller.

## Collapse stratification dimensions

The filtered cube still contains other demographic dimensions. To
compare total population by county and year, collapse the remaining
stratification dimensions to `"All"`. For this cube, that means
collapsing `sex`, `age.char`, `race`, and `ethnicity`.

```{r  collapse_pop_small}
pop_totals <- pop_small |>
  collapse_dim(
    dim = "sex",
    groups = stats::setNames(
      rep("All", length(dimnames(pop_small)$sex)),
      dimnames(pop_small)$sex
    )
  ) |>
  collapse_dim(
    dim = "age.char",
    groups = stats::setNames(
      rep("All", length(dimnames(pop_small)$age.char)),
      dimnames(pop_small)$age.char
    ),
    allow_overlap = TRUE
  ) |>
  collapse_dim(
    dim = "race.eth",
    strict = FALSE,
    groups = stats::setNames(
      rep("All", length(dimnames(pop_small)$race.eth )),
      dimnames(pop_small)$race.eth
    )
  ) 

pop_totals
```

## Convert to a tibble

Converting to a tibble is an eager step, so it is best done after
filtering and collapsing.

```{r}
county_df <- pop_totals |>
  as_tibble() |>
  dplyr::mutate(year = as.integer(as.character(year)))

county_df
```

## Plot the result

```{r}
#| fig.width: 8
#| fig.height: 4.8
ggplot(county_df, aes(x = year, y = population, color = area.name)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Population comparison for three Texas counties",
    subtitle = "Recent years from tdc_estimates_county",
    x = "Year",
    y = "Population",
    color = "County"
  ) +
  theme_minimal(base_size = 12)
```

# What to remember

The `tarr.pop` package provides `poparray` as one of its core classes.

That class gives you:

-   a cube-shaped representation of population data
-   lazy computation through DelayedArray
-   disk-backed storage through HDF5Array
-   semantic metadata through `DimSemantics`

Taken together, those pieces make it easier to work with large
population data in a way that is both scalable and easier to reason
about.

# Next steps

After you are comfortable opening and filtering cubes, common next steps
are:

-   exploring `dim_semantics()` in more detail
-   using `collapse_dim()` for summaries
-   converting a filtered cube to a tibble for reporting
-   using `project()` to create future population projections
