---
title: "Introduction to tarr.pop"
format: html
vignette: >
  %\VignetteIndexEntry{"Introduction to tarr.pop"}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
library(tarr.pop)
library(dplyr)
library(tibble)
library(ggplot2)
```

# Introduction

The `tarr.pop` package is designed for working with population denominators as
multi-dimensional data rather than as long tables of rows. That distinction is
important in epidemiology and demography, where a population count is usually
defined by several dimensions at once:

- year
- geography
- age
- sex
- race
- ethnicity

In a regular table, those dimensions are present, but their structure is only
implied by columns. In `tarr.pop`, the same data is treated as a population
cube, which makes the structure explicit and easier to work with safely.

# Why use a population cube?

Population data is naturally multi-dimensional. A single cell in a cube might
mean:

> the number of people in Tarrant County, in 2022, age 65 to 69, female

That is not a row-first idea. It is a combination of labeled dimensions.

The `tarr.pop` package uses the `poparray` class to represent this structure.
`poparray` builds on [DelayedArray] and [HDF5Array], so cubes can stay on disk
and be manipulated lazily until you actually need a realized result.

# The `poparray` class

A `poparray` is an S4 class that stores:

- the population values themselves
- the names and labels of each dimension
- the role of key dimensions such as time and area
- semantic information about each dimension

This means a `poparray` is more than an array of numbers. It is an array plus a
contract describing what those numbers mean.

## Roles

Every `poparray` has:

- one time dimension
- one area dimension
- zero or more additional stratification dimensions

Those roles are stored with the object, so functions can work with the cube by
meaning rather than by hard-coded position.

## `DimSemantics`

The `DimSemantics` class stores semantic information for each dimension. For
example, it can describe whether a dimension behaves like:

- a partition, where categories are mutually exclusive
- a set, where categories may overlap
- an interval scale, such as age groups or years

That matters because not every aggregation is safe. Some categories overlap and
should not be summed without care. By storing semantics with the cube,
`tarr.pop` can help prevent invalid operations.

You can inspect the semantic metadata directly:

```{r}
pop <- open_poparray("tdc_estimates_county")

names(dim_semantics(pop))
dim_semantics(pop)$year
dim_semantics(pop)$age.char
```

# A first workflow

This section walks through a simple beginner workflow:

1. open a cube
2. inspect the structure
3. filter to a smaller analysis slice
4. collapse remaining stratification dimensions
5. convert to a tibble for plotting

## Open a cube

For this example, use the Texas Demographic Center county estimates cube.

```{r}
pop <- open_poparray("tdc_estimates_county")
pop
```

## Inspect the structure

Start by asking a few basic questions about the cube.

```{r}
dim(pop)
names(dimnames(pop))
time_role(pop)
area_role(pop)
```

The dimension labels are also easy to inspect.

```{r}
head(dimnames(pop)[[time_role(pop)]])
head(dimnames(pop)[[area_role(pop)]])
```

## Filter to a smaller slice

Suppose we want to compare Dallas, Tarrant, and Travis counties over the most
recent five years in the cube.

```{r}
target_counties <- c("Dallas", "Tarrant", "Travis")

recent_years <- dimnames(pop)[[time_role(pop)]] |>
  tail(5)

pop_small <- pop |>
  dplyr::filter(
    area.name %in% target_counties,
    year %in% recent_years
  )

pop_small
```

This filtering step is still lazy. At this point, `tarr.pop` has created a
smaller cube view, but it has not converted the full result into an in-memory
data frame.

## Collapse stratification dimensions

The filtered cube still contains other demographic dimensions. To compare total
population by county and year, collapse the remaining stratification dimensions
to `"All"`.

```{r}
strata_dims <- names(dimnames(pop_small)) |>
  setdiff(c(time_role(pop_small), area_role(pop_small)))

pop_totals <- pop_small

for (dm in strata_dims) {
  labs <- dimnames(pop_totals)[[dm]]
  groups <- stats::setNames(rep("All", length(labs)), labs)
  pop_totals <- collapse_dim(pop_totals, dim = dm, groups = groups)
}

pop_totals
```

## Convert to a tibble

Converting to a tibble is an eager step, so it is best done after filtering and
collapsing.

```{r}
county_df <- pop_totals |>
  as_tibble() |>
  dplyr::mutate(year = as.integer(as.character(year)))

county_df
```

## Plot the result

```{r}
#| fig.width: 8
#| fig.height: 4.8
ggplot(county_df, aes(x = year, y = population, color = area.name)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Population comparison for three Texas counties",
    subtitle = "Recent years from tdc_estimates_county",
    x = "Year",
    y = "Population",
    color = "County"
  ) +
  theme_minimal(base_size = 12)
```

# What to remember

The `tarr.pop` package is the package. `poparray` is one of its core classes.

That class gives you:

- a cube-shaped representation of population data
- lazy computation through DelayedArray
- disk-backed storage through HDF5Array
- semantic metadata through `DimSemantics`

Taken together, those pieces make it easier to work with large population
denominators in a way that is both scalable and easier to reason about.

# Next steps

After you are comfortable opening and filtering cubes, common next steps are:

- exploring `dim_semantics()` in more detail
- using `collapse_dim()` for summaries
- converting a filtered cube to a tibble for reporting
- using `project()` to create future population projections
