---
title: "Introduction to poparray: Compare Dallas, Tarrant, and Travis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to poparray: Compare Dallas, Tarrant, and Travis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Goal

This vignette introduces the `poparray` workflow by answering one concrete question:

**How do total populations compare over time for Dallas, Tarrant, and Travis counties?**

You will:

1. Open a county-level population cube.
2. Inspect dimensions and semantic roles.
3. Filter to three counties and recent years.
4. Collapse strata dimensions to county totals.
5. Plot the comparison with `ggplot2`.

```{r setup}
library(tarr.pop)
library(dplyr)
library(tibble)
library(ggplot2)
```

## 1) Choose and open a population cube

For this example, use the county-level Texas Demographic Center estimates:

```{r open-cube}
series_id <- "tdc_estimates_county"
pop <- open_poparray(series_id)
pop
```

## 2) Inspect the cube

Check dimensions and roles so you know which axes to manipulate.

```{r inspect-cube}
dim(pop)
names(dimnames(pop))
time_role(pop)
area_role(pop)
```

Confirm the three counties are present:

```{r check-counties}
target_counties <- c("Dallas", "Tarrant", "Travis")
all(target_counties %in% dimnames(pop)[[area_role(pop)]])
```

## 3) Filter to the analysis subset

Use all available years in the cube, then keep only the most recent five years.
This avoids hard-coding year labels.

```{r filter-subset}
year_labels <- as.integer(as.character(dimnames(pop)[[time_role(pop)]]))
recent_years <- sort(year_labels, decreasing = TRUE)[1:5] |> sort()

pop_sel <- dplyr::filter(
  pop,
  area.name %in% target_counties,
  year %in% recent_years
)

pop_sel
```

## 4) Collapse strata to county totals

`pop_sel` still contains stratification dimensions (for example sex, age, and race/ethnicity).
Collapse each non-time/non-area dimension to one `"All"` group so each county-year has one total value.

```{r collapse-totals}
strata_dims <- setdiff(
  names(dimnames(pop_sel)),
  c(time_role(pop_sel), area_role(pop_sel))
)

pop_totals <- pop_sel
for (dm in strata_dims) {
  labs <- dimnames(pop_totals)[[dm]]
  groups <- stats::setNames(rep("All", length(labs)), labs)
  pop_totals <- collapse_dim(pop_totals, dim = dm, groups = groups)
}

pop_totals
```

## 5) Convert to a tibble for plotting

Conversion to a tibble is eager, so it is best done after filtering/collapsing.

```{r to-tibble}
county_df <- as_tibble(pop_totals) |>
  dplyr::mutate(year = as.integer(as.character(year)))

county_df
```

## 6) Plot county trends with ggplot2

```{r plot-county-compare, fig.width=8, fig.height=4.8}
ggplot(county_df, aes(x = year, y = population, color = area.name)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Population Comparison: Dallas, Tarrant, and Travis",
    subtitle = "Recent years from tdc_estimates_county",
    x = "Year",
    y = "Population",
    color = "County"
  ) +
  theme_minimal(base_size = 12)
```

## 7) Optional summary table

```{r summary-table}
county_summary <- county_df |>
  dplyr::group_by(area.name) |>
  dplyr::arrange(year, .by_group = TRUE) |>
  dplyr::summarise(
    first_year = dplyr::first(year),
    last_year = dplyr::last(year),
    first_pop = dplyr::first(population),
    last_pop = dplyr::last(population),
    absolute_change = last_pop - first_pop,
    pct_change = 100 * (last_pop / first_pop - 1),
    .groups = "drop"
  )

county_summary
```

## Next steps

- Use `project()` to forecast future populations from a filtered `poparray`.
- Explore other county cubes (for example Census or SEER series) with the same workflow.
- Build rate denominators by joining county totals to case/event data.
