---
title: "Projection Workflow with poparray: Dallas, Tarrant, and Travis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Projection Workflow with poparray: Dallas, Tarrant, and Travis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Goal

This vignette extends the introductory workflow by projecting forward county totals for:

- Dallas
- Tarrant
- Travis

You will:

1. Build a county-total `poparray` for recent base years.
2. Run `project()` to create a `poparray_projection`.
3. Extract projected values and uncertainty.
4. Plot historical and projected county trends together.

```{r setup}
library(tarr.pop)
library(dplyr)
library(tibble)
library(ggplot2)
```

## 1) Open a county-level cube and define the analysis subset

```{r open-and-filter}
pop <- open_poparray("tdc_estimates_county")

target_counties <- c("Dallas", "Tarrant", "Travis")

year_labels <- dim_labels(pop, "year") |> as.integer()
base_years <- sort(year_labels, decreasing = TRUE)[1:7] |> sort()

pop_base <- dplyr::filter(
  pop,
  area.name %in% target_counties,
  year %in% base_years
)

pop_base
```

## 2) Collapse strata to county totals

`project()` can run on stratified cubes, but for this example we project total county population.  To do so,
collapse each of the dimensions that are not year or area to 'All'.  When collapsing like this, all of 
the dimension labels (e.g. `female` and `male` for `sex`) are summed and 'All' is the new label for sex. If one of the 
dimensions has labels or groups that overalp with another entry in that dimension, the sum will be inaccurate, 
therefore the `collapse_dim()` function will stop with an error when unsafe sum/aggregation may occur.  There may  times that the "unsafe" behavior is desired. To allow  Set `strict` argument to `FALSE` and the `collapse_dim()` function will issue a warning but still allow the sum to happen.

```{r collapse-to-totals}
# Get the dimensions that are not the time and area roles
strata_dims <- setdiff(
  names(dimnames(pop_base)),
  c(time_role(pop_base), area_role(pop_base))
)

pop_totals <- pop_base

# reduce the po_base to have all in the identified dimensions
# run through each dimension and collapse
pop_totals <- purrr::reduce(strata_dims, 
                     function(acc, dm){
                       labs <- dim_labels(acc, dm)
                       new_group <- list("All" = labs)
                       collapse_dim(acc, dim = dm, groups = new_group)
                     }, 
                     .init = pop_base
)

# for (dm in strata_dims) {
#   labs <- dim_labels(pop_totals, dm)
#   groups <- list("All" = labs)
#   pop_totals <- collapse_dim(pop_totals, dim = dm, groups = groups)
# }

pop_totals
```

## 3) Run a projection

Project 5 years ahead with automatic method selection and 95% intervals.

```{r run-projection}
pr <- project(
  pop_totals,
  h = 5,
  method = "auto",
  level = 0.95,
  guard = FALSE
)

show(pr)
```

## 4) Convert projected output to a tibble

Use `include_confidence = TRUE` to include lower/upper bounds.

```{r projection-table}
proj_df <- as_tibble(pr, include_confidence = TRUE) |>
  dplyr::mutate(year = as.integer(year))

proj_df
```

## 5) Build a historical + projected comparison frame

```{r combine-historical-projected}
hist_df <- as_tibble(pop_totals) |>
  dplyr::transmute(
    year = as.integer(as.character(year)),
    area.name,
    value = population,
    series = "Historical"
  )

proj_line_df <- proj_df |>
  dplyr::transmute(
    year,
    area.name,
    value = projection,
    series = "Projected",
    lower,
    upper
  )

plot_df <- dplyr::bind_rows(
  hist_df,
  dplyr::select(proj_line_df, year, area.name, value, series)
)
```

## 6) Plot county trends with projection intervals

```{r plot-projection-compare, fig.width=9, fig.height=5.2}
ggplot() +
  geom_ribbon(
    data = proj_line_df,
    aes(x = year, ymin = lower, ymax = upper, fill = area.name),
    alpha = 0.14,
    color = NA
  ) +
  geom_line(
    data = plot_df,
    aes(x = year, y = value, color = area.name, linetype = series),
    linewidth = 1.05
  ) +
  geom_point(
    data = plot_df,
    aes(x = year, y = value, color = area.name, shape = series),
    size = 1.8
  ) +
  scale_y_continuous(labels = scales::label_comma()) +
  labs(
    title = "Historical and Projected Population: Dallas, Tarrant, Travis",
    subtitle = "Shaded ribbons show 95% projection intervals",
    x = "Year",
    y = "Population",
    color = "County",
    fill = "County",
    linetype = "Series",
    shape = "Series"
  ) +
  theme_minimal(base_size = 12)
```

## 7) Quick projected summary

```{r projected-summary}
proj_summary <- proj_df |>
  dplyr::group_by(area.name) |>
  dplyr::arrange(year, .by_group = TRUE) |>
  dplyr::summarise(
    first_projection_year = dplyr::first(year),
    last_projection_year = dplyr::last(year),
    first_projection = dplyr::first(projection),
    last_projection = dplyr::last(projection),
    projection_change = last_projection - first_projection,
    .groups = "drop"
  )

proj_summary
```

## Next steps

- Compare methods explicitly (`ETS`, `CAGR`, `ARIMA`) for the same base cube.
- Keep selected strata (for example sex or age) and project subgroup trajectories.
- Export `as_tibble(pr, include_confidence = TRUE)` for reporting pipelines.
