---
title: "The poparray class and objects"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The poparray class and objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tarr.pop)
```

# poparray: A Lazy, Role-Aware Population Array

## Overview

A **`poparray`** object represents population counts indexed over **time** and **area**, optionally stratified by additional dimensions such as age, sex, race, household size, or economic characteristics.

The class is designed for **large, multi-dimensional population data**, with an emphasis on:

-   scalability (HDF5-backed storage),
-   delayed computation,
-   explicit dimensional contracts, and
-   clear separation between numeric data and semantic metadata.

`poparray` is intended for epidemiologic analysis, demographic reporting, and population projection workflows where correctness, traceability, and memory safety are critical.

------------------------------------------------------------------------

## Core design principles

The design of `poparray` is governed by the following principles:

1.  **Time and area are structural invariants**\
    Every `poparray` must include:

    -   exactly one **time dimension** (e.g. `year`), and
    -   exactly one **area dimension** (e.g. `area`, `county`, `tract`).

2.  **All other dimensions are optional stratifications**\
    Additional dimensions represent orthogonal population stratifications and may include:

    -   demographic variables (age, sex, race, ethnicity),
    -   socioeconomic variables (income bracket, education level),
    -   household characteristics (household size, family type),
    -   or other analyst-defined groupings.

3.  **Only independent indexing axes are dimensions**\
    Hierarchical geography (e.g. tract → county → state) is **not** encoded as multiple array dimensions.\
    Instead, only one geographic resolution is represented in the array at any time.

4.  **Hierarchy and meaning live in metadata, not in the array**\
    Lookup tables and provenance information are stored as attributes, not as dimensions or additional array axes.

5.  **Delayed computation by default**\
    Operations preserve laziness whenever possible and avoid realizing large arrays in memory.

------------------------------------------------------------------------

## What a poparray object contains

A `poparray` object is an S3 wrapper around a multi-dimensional array (often HDF5-backed) plus semantic metadata.

### Numeric data

-   Population counts stored as:
    -   a base R `array`, or
    -   a `DelayedArray` / `HDF5Array` for large datasets.
-   Dimension names (`dimnames`) are always present and meaningful.
-   Chunking and on-disk storage are handled by the HDF5Array infrastructure.

### Required dimensions

Every `poparray` includes:

| Role | Description |
|----------------------|--------------------------------------------------|
| Time | An ordered time dimension (e.g. `year`) |
| Area | A single geographic indexing dimension (e.g. `block`, `tract`, `county`) |

Dimension order is not significant; semantic roles are determined by metadata.

### Optional stratification dimensions

Any number of additional dimensions may be present, representing orthogonal population stratifications.\
These dimensions:

-   are independent of time and area,
-   are not hierarchical with each other,
-   may be collapsed or retained depending on analysis needs.

------------------------------------------------------------------------

## Attribute usage and semantic metadata

`poparray` relies heavily on attributes to store meaning that should not be encoded structurally in the array.

## dim_roles

A named list describing the semantic role of each dimension:

`attr(x, "dim_roles") <- list( time = "year",     area = "block",     strata = c("age", "sex") )`

This allows methods to operate based on role, rather than hard-coded dimension names or positions.

area_lookup A data frame describing hierarchical geographic relationships for the area dimension.

Example:

`attr(x, "area_lookup") \<- data.frame( block = "...", tract = "...", county = "...", state = "TX" )` 

This table:

has one row per area level in the array,

encodes many-to-one relationships,

is used for aggregation and reconciliation,

is never treated as an array dimension.

Source and provenance metadata Attributes such as:

source

series_spec

release year

Census vintage

differential privacy notes

These document where the data came from and how it should be interpreted, without affecting array mechanics.

Storage and persistence model A poparray object typically consists of two files:

HDF5 file Stores the numeric array data.

Handles chunking, compression, and on-disk access.

Contains no semantic metadata beyond dimnames.

RDS file Stores the S3 poparray wrapper.

Contains all attributes, including lookup tables and provenance.

References (but does not embed) the HDF5 file.

This separation allows:

multiple derived views of the same numeric data,

safe sharing across users,

metadata evolution without rewriting large arrays.

Supported operations Subsetting ([) Preserves laziness.

Maintains dimensional integrity.

Updates metadata consistently.

Defaults to drop = FALSE.

Collapsing dimensions Stratification dimensions may be collapsed freely.

Area and time dimensions may be collapsed only explicitly, with clear intent.

Area collapsing uses lookup tables and replaces (not adds) the area dimension.

Geographic aggregation Fine-to-coarse aggregation (e.g. block → county) is supported via lookup-driven collapsing.

The resulting object remains a valid poparray.

Benchmark reconciliation (optional) When fine-level data are subject to differential privacy noise:

Aggregated totals may differ from authoritative higher-level totals.

Reconciliation methods (e.g. benchmarking or raking) may be applied explicitly.

Provenance of reconciliation is recorded in metadata.

Plotting and summaries Methods may impose constraints on remaining stratification dimensions.

All plotting and summaries respect dimension roles and metadata.

What poparray deliberately does not do It does not encode hierarchical geography as multiple dimensions.

It does not assume a fixed set of demographic variables.

It does not silently reconcile or adjust data.

It does not force realization of large arrays without warning.

Intended use cases Census and intercensal population analysis

Epidemiologic rate calculations

Population projections

Small-area estimation workflows

Shared, reproducible population cubes across teams

Conceptual summary A poparray is a lazy, role-aware population cube where arrays store values, metadata stores meaning, and hierarchy is expressed through lookup tables, not dimensions.
