---
title: "Creating a Population Cube"
format: html
vignette: >
  %\VignetteIndexEntry{Creating a Population Cube}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
execute:
  eval: false
editor:
  markdown:
    wrap: 100
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

This vignette shows an example of how a population cube can be created from Texas Demographic Center
estimate CSV files. The example follows the workflow in `data-raw/tdc_estimates.r`.

The CSV files were sourced

The purpose is not just to read files. The goal is to move from source-specific, wide-format CSV
files into a `poparray` cube with:

-   named dimensions
-   explicit time and area roles
-   dimension semantics
-   HDF5-backed storage
-   source metadata

The source CSV files are local raw data, so this vignette is written as a reproducible template
rather than code that runs during package checks. The resulting cube is written to the active cube
store, not into the package source tree, unless the cube path has been configured that way.

# Prerequisites

The raw-data script assumes the same setup used by `data-raw/control_def.r`: supporting packages are
loaded, package development helpers are available, and the source-data paths are defined.

For an interactive development session, the setup looks like this:

```{r setup}
library(tidyverse)
library(janitor)
library(data.table)
library(rage)
library(DelayedArray)

devtools::load_all()

paths <- tarr::paths
```

The estimate CSV files are expected under the population source folder:

```{r source-folder}
input_dir <- file.path(
  tarr::paths$population,
  "Estimates",
  "Texas Demographic Center",
  "asre"
)
```

The file names are expected to begin with a four-digit year and end with
`_ASRE_Estimate_alldata.csv`, for example:

``` text
2017_ASRE_Estimate_alldata.csv
2018_ASRE_Estimate_alldata.csv
```

# Workflow

The cube creation workflow has five stages:

1.  Define source-specific helper functions.
2.  Read the raw CSV files into one table.
3.  Transform the table into canonical cube columns.
4.  Define dimension semantics and a support table.
5.  Ingest the table into an HDF5-backed `poparray` cube.

The first three stages are source-specific. They know about Texas Demographic Center column names
and labels. The last stage uses `ingest_population()`, which is the package-level ingestion function
for validating and writing a cube.

# Clean Age Labels

TDC age labels are not always in the same form that `tarr.pop` should store. For example, labels may
include words such as `"Ages"` or `"years"`, and open-ended age groups may need spacing normalized.

`process_age_char()` standardizes those labels before they become a cube dimension.

```{r process-age}
process_age_char <- function(x) {
  x |>
    stringr::str_remove_all(stringr::regex("Ages", ignore_case = TRUE)) |>
    stringr::str_remove_all(stringr::regex(" (ye?a??rs?|Ages)", ignore_case = TRUE)) |>
    stringr::str_trim(side = "both") |>
    stringr::str_replace("5\\+", "5 +")
}
```

Age is not treated as a plain character field. In a population cube, age labels have interval
meaning. The helper below uses `rage::as.age_group()` to sort age groups by their interval order,
not by alphabetic order.

```{r ordered-age-levels}
ordered_age_levels <- function(x) {
  age_levels <- levels(x)
  age_levels <- age_levels[age_levels != "All"]
  c(sort(as.character(rage::as.age_group(age_levels))), "All")
}
```

The `"All"` level is kept temporarily while source labels are being normalized. It is removed before
the data are ingested because aggregate rows should not be physically stored in a `poparray` cube.

# Read One CSV

The CSV files are read with explicit column types. County, FIPS, and age are read as factors because
their labels define dimensions. Most population columns are integer counts.

`read_est_csv()` also extracts the estimate year from the file name. That makes the year a data
column even when the file itself does not include one.

```{r read-one-csv}
read_est_csv <- function(file, col_types) {
  readr::read_csv(
    file = file,
    col_types = col_types,
    id = "file_name",
    progress = TRUE
  ) |>
    janitor::clean_names() |>
    dplyr::mutate(
      year = basename(file_name) |>
        stringr::str_extract("^20[1-2][0-9]") |>
        as.integer()
    ) |>
    dplyr::select(-file_name) |>
    dplyr::rename_with(\(col) {
      col |>
        stringr::str_replace("anglo", "white") |>
        stringr::str_remove("^nh_")
    })
}
```

The name cleanup is source-specific. TDC labels changed over time, and this step normalizes older
and newer naming so the final cube has one consistent `race.eth` dimension.

# Read All Estimate Files

`read_tdc_estimates_raw()` finds all matching CSV files in the source directory and uses
`read_est_csv()` on each one. The separate files are then bound into one `data.table`.

```{r read-all-csvs}
read_tdc_estimates_raw <- function(
    pattern = "20[1-2][0-9]_ASRE_Estimate_alldata\\.csv",
    input_dir = file.path(tarr::paths$population, "Estimates", "Texas Demographic Center", "asre"),
    ...
) {
  files <- list.files(path = input_dir, pattern = pattern, full.names = TRUE)

  if (!length(files)) {
    cli::cli_abort("No TDC estimate files matched {.val {pattern}} in {.file {input_dir}}.")
  }

  col_types <- readr::cols(
    County = readr::col_factor(),
    FIPS = readr::col_factor(),
    Age = readr::col_factor(ordered = TRUE),
    .default = readr::col_integer()
  )

  purrr::map(files, read_est_csv, col_types = col_types) |>
    dplyr::bind_rows() |>
    data.table::setDT()
}
```

At this point the data are still close to the source layout. The demographic counts are still spread
across many columns.

# Transform to Cube Columns

The transformer is the most important source-specific part of the workflow. It turns the wide TDC
layout into the canonical long table expected by `ingest_population()`.

The canonical columns for this cube are:

```{r dims}
dims <- c("year", "area.name", "sex", "age.char", "race.eth")
```

Together with the value column `population`, each row describes one cell in the eventual cube.

```{r transform}
transform_tdc_estimates <- function(df, counties = NULL, include_texas_total = FALSE) {
  stopifnot(data.table::is.data.table(df))

  wide_names <- names(df) |>
    stringr::str_replace_all("_", ".") |>
    stringr::str_replace("^total$", "All.All") |>
    stringr::str_replace("^total", "All") |>
    stringr::str_replace("\\.total$", ".All")

  data.table::setnames(df, wide_names)

  long <- data.table::melt(
    data = df,
    id.vars = c("year", "fips", "county", "age"),
    variable.name = "race.sex",
    variable.factor = TRUE,
    value.name = "population",
    verbose = FALSE
  )

  long[
    ,
    c("race.eth", "sex") := data.table::tstrsplit(race.sex, split = ".", fixed = TRUE, fill = NA)
  ][
    ,
    county := forcats::fct_relabel(
      county,
      \(x) stringr::str_to_title(gsub(" COUNTY", "", x, ignore.case = TRUE))
    )
  ][
    ,
    county := forcats::fct_recode(county, "Texas" = "State Of Texas")
  ][
    ,
    age := forcats::fct_relabel(
      age,
      \(x) process_age_char(x) |> rage::as.age_group() |> as.character()
    )
  ][
    ,
    age := ordered(age, levels = ordered_age_levels(age))
  ][
    ,
    sex := forcats::fct_na_value_to_level(sex, level = "All")
  ][
    ,
    race.eth := factor(race.eth)
  ][
    ,
    c("race.sex", "fips") := NULL
  ]

  data.table::setnames(long, old = c("age", "county"), new = c("age.char", "area.name"))

  if (!is.null(counties)) {
    long <- long[area.name %chin% counties]
  }

  if (!isTRUE(include_texas_total)) {
    long <- long[area.name != "Texas"]
  }

  long <- long[!is.na(population)]

  dim_cols <- c("area.name", "sex", "age.char", "race.eth")
  keep <- long[
    ,
    !Reduce(`|`, lapply(.SD, \(x) as.character(x) == "All")),
    .SDcols = dim_cols
  ]
  long <- long[keep]

  data.table::setorder(long, year, area.name)

  factor_cols <- names(long)[vapply(long, is.factor, logical(1))]
  if (length(factor_cols)) {
    long[, (factor_cols) := lapply(.SD, droplevels), .SDcols = factor_cols]
  }

  data.table::setcolorder(
    long,
    c("year", "area.name", "sex", "age.char", "race.eth", "population")
  )

  long
}
```

This function does several jobs:

The first block normalizes the source column names. Columns such as totals are renamed to make the
implicit race and sex labels explicit. This is needed before the wide table is melted.

The `melt()` call changes the data from wide format to long format. After that operation, the
combined source column name is split into `race.eth` and `sex`.

County names are normalized to title case and `"State Of Texas"` is recoded to `"Texas"`. The state
total is then removed by default. This matters because a county cube should have an `area.name`
dimension that is a true partition. If the state total is stored together with counties, the area
dimension would overlap.

Rows containing `"All"` are removed before ingestion. In `tarr.pop`, `"All"` is treated as a virtual
aggregate level, not a physical level to store in a base cube. Aggregate rows can be recomputed
later by grouping or collapsing dimensions.

The final `droplevels()` call removes factor levels that are no longer present after filtering. This
is especially important before creating a support table, because stale factor levels such as `"All"`
or `"Texas"` can otherwise be carried into the support skeleton.

# Define Dimension Semantics

Every cube needs semantics for its dimensions. These semantics describe what the dimension means and
how it can safely be used.

```{r semantics}
tdc_estimate_semantics <- function() {
  list(
    year = tarr.pop:::new_dim_semantics(
      dim_name = "year",
      domain = "time",
      partition_type = "partition",
      scale_type = "interval",
      validated = TRUE,
      notes = "July 1 estimate year."
    ),
    area.name = tarr.pop:::new_dim_semantics(
      dim_name = "area.name",
      domain = "area",
      partition_type = "partition",
      scale_type = "nominal",
      validated = TRUE,
      notes = "County-only geography. Texas state total is excluded from this cube."
    ),
    sex = tarr.pop:::new_dim_semantics(
      dim_name = "sex",
      domain = "sex",
      partition_type = "partition",
      scale_type = "nominal",
      validated = TRUE
    ),
    age.char = tarr.pop:::new_dim_semantics(
      dim_name = "age.char",
      domain = "age interval",
      partition_type = "set",
      scale_type = "interval",
      validated = TRUE,
      overlap_levels = "85 +",
      notes = "Age groups are interval-valued and may include an open upper bound."
    ),
    race.eth = tarr.pop:::new_dim_semantics(
      dim_name = "race.eth",
      domain = "race and ethnicity",
      partition_type = "partition",
      scale_type = "nominal",
      validated = TRUE,
      notes = paste(
        "TDC combines race and Hispanic ethnicity into one dimension.",
        "The Hispanic category is treated as a level in this partition."
      )
    )
  )
}
```

The time and area dimensions are marked as partitions. That is required by the `poparray`
constructor because those roles must identify mutually exclusive cells.

The age dimension is marked as interval-valued. Even though age labels are stored as strings, they
represent intervals such as one-year ages and open-ended groups.

# Create a Support Table

The support table lists valid dimension combinations that should exist in the cube even when the
source table does not contain an observed row. In this TDC example, the source did not include
`asian` race and ethnicity estimates before 2017. If those cells should exist in the cube as missing
values, the support table supplies the valid skeleton and `completion_policy = "na"` fills them with
`NA`.

Support tables should use the same dimension names and compatible classes as the transformed table.
This is particularly important for ordered factors such as `age.char`, because their order
determines the final cube layout.

```{r support-table}
default_counties <- NULL
if (exists("county_fips", inherits = TRUE)) {
  default_counties <- setdiff(names(county_fips), "Texas")
}

age_levels <- c(
  "< 1", as.character(1:84), "85", "85 +",
  as.character(86:94), "95 +"
) |>
  rage::as.age_group() |>
  ordered()

support_table <- tidyr::expand_grid(
  year = 2011:2016,
  area.name = default_counties,
  sex = c("male", "female"),
  age.char = age_levels,
  race.eth = "asian"
)
```

Use `completion_policy = "na"` for this case. The missing values are not zero population. They
represent estimates that are valid for the harmonized cube schema but unavailable in the source
files.

# Ingest the Cube

`ingest_population()` is the package-native ingestion entry point. It runs the reader, runs the
transformer, removes aggregate rows when requested, applies the completion policy, validates the
table, converts it to an array, and writes the HDF5-backed cube.

```{r ingest}
cube_root <- tarr.pop::init_cubes()
tdc_estimates_file <- file.path(cube_root, "base", "tdc_estimates_county.h5")

tarr.pop::ingest_population(
  reader = read_tdc_estimates_raw,
  transformer = transform_tdc_estimates,
  dims = dims,
  dim_semantics = tdc_estimate_semantics(),
  filepath = tdc_estimates_file,
  series_id = "tdc_estimates_county",
  completion_policy = "na",
  drop_all = TRUE,
  source_meta = list(
    note = "Texas Demographic Center county estimates",
    population_type = "Estimate",
    source = "Texas Demographic Center, Estimates program"
  ),
  time_dim = "year",
  area_dim = "area.name",
  support = support_table,
  overwrite = TRUE,
  data_col = "population",
  counties = default_counties,
  include_texas_total = FALSE
)
```

The output file path is:

```{r output-path}
tdc_estimates_file
```

This path is under `file.path(tarr.pop::cube_path(), "base")`. The cube path is user-configurable
with `set_cube_path()`.

# Open and Verify the Cube

After ingestion, reopen the cube by its series identifier.

```{r verify}
est_poparray <- tarr.pop::open_poparray("tdc_estimates_county")
est_poparray

dim(est_poparray)
names(dimnames(est_poparray))
tarr.pop::time_role(est_poparray)
tarr.pop::area_role(est_poparray)
```

The object is a `poparray` backed by HDF5. Opening it is lazy: the full cube is not read into
memory. Values are read only when a later operation requires them.

Useful checks after creating a cube include:

```{r verify-labels}
tarr.pop::years(est_poparray)
tarr.pop::ages(est_poparray)
tarr.pop::dim_semantics(est_poparray)
tarr.pop::get_source(est_poparray)
```

# Common Problems

A support table with mismatched classes can produce joins that appear to lose observed values. Build
support from the same labels and factor order used in the transformed table, especially for
`age.char`.

Rows containing `"All"` should be removed before ingestion. The base cube should store detailed
cells, not aggregate rows. Aggregate `"All"` results can be created later from the cube.

The Texas state total should not be stored in the county cube. Keeping `"Texas"` together with
counties makes `area.name` overlap, which violates the partition semantics expected for the area
role.

Converting a cube to a data frame or tibble is **EAGER** because it realizes cube values into rows.
Creation and opening of the HDF5-backed cube are designed so that the resulting `poparray` remains
lazy for downstream analysis.
