---
title: "Creating and Verifying a Population Cube"
subtitle: "A guide using Texas Demographic Center county estimates as an example"
format:
  html:
    toc: true
    toc-depth: 3
execute:
  cache: false
  eval: false
vignette: >
  %\VignetteIndexEntry{Creating and Verifying a Population Cube}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
editor:
  markdown:
    wrap: 100
---

# Purpose and scope

A poparray object created by this package is stored on disk in a population "cube" or array  that has a count at each
combination of demographic dimensions.  There are two required dimensions that can referred to by their roles: "time" and
"area".  Any data set that has time and area with population counts can be made into a poparray and saved as a population
cube. Most population cubes built for this package have year, area.name, age.char, sex, race, and ethnicity dimensions.  
However a poparray can be built from other data such as fields calleddecade, census_tract, house_age_range, and 
population values. 

For this vignette, the annual Texas Demographic Center (TDC) county estimate files are used as an example of how a
poparray and population cube are created. The data files are saved in comma separated value (csv) text format with a
file for each year from 2011 through 2024. The column in the files have county, FIPS/Area Code, Sex, Age, and 
Race Ethnicity (NH-Asian, NH-Black, NH-White, Hispanic, etc) with population figures for each combination of variables.
The TDC has changed file schema over the years.  For example, what used to be "Anglo" is now NH-White.  The "Other" race
category included Asians until 2017 when that became separate category. Later files also include the "year" of the
estimate in a column wile earlier files do not have that column.  Files before 2017 use "FIPS" for the area code, where
as later files use the term "Area Code" but contain the same data as the older "FIPS". Finally, the estimates use
single ages and in files before 2017 that was 0 through 85+.  After 2017 the ages listed were 0 to 95+.

Because the file columns and their names as well as some category values changed over the years, steps are added in
constructing the population cube.  In addition to renaming columns to standard names, the lack of an Asian race category
in files before 2017, means population figures for that category are missing in those earlier years. How this is handled
and what it means for a person using this poparray are covered later in this vignette.  Finally, the earlier year TDC
estimates include single ages from 0 (infant) thorough 84 and 85+ as the last group.  Later files included ages 0
through 94 with 95+ being the last group.

After reading each file, the resulting data frames are standardized to use the same column names then transformed into
columns that will match the eventual array dimensions. Those dimension names are year (the time role), county (the area
role), sex, age.char intervals, and combined race/ethnicity. The finished object is a `poparray` whose values reside in
an HDF5 file and whose labels, dimension roles, semantics, and provenance are stored with those values.

Creating a trustworthy cube requires more than a successful file write. Source identities must survive transformation,
every observed value must keep its coordinate, and missing values must be classified correctly. This guide follows the
current repository workflow, builds to an isolated location first, verifies the result, and only then describes
replacing a production cube.

The examples here require a source checkout of `tarr.pop` and local annual CSV files. Execution is disabled throughout
the document so rendering it cannot launch an ingestion job or overwrite a cube. Run the chunks deliberately, in order,
in a fresh R session from the repository root.

# Understand the data model first

## Dimensions and roles

The canonical table has exactly these columns, in this order:

| Column | Meaning | Expected representation |
|---|---|---|
| `year` | Annual population snapshot | Integer year |
| `area.name` | Texas county name | Factor |
| `sex` | female or male | Factor |
| `age.char` | Age interval label | Ordered factor |
| `race.eth` | Combined race/ethnicity category | Factor |
| `population` | Population count | Numeric or integer, and  `NA` for missing|

The time role points to `year`; the area role points to `area.name`. Roles identify the relevant dimensions without
relying on their positions. Dimension semantics describe which operations are valid, including whether the labels form a
disjoint partition or a potentially overlapping set.

## TDC age and race/ethnicity support

When creating the poparray, all combinations of the dimensions should be represented in a "support table".  With the
changes that TDC made to the schema in files for 2017 and later, the support table must have the name/category combinations
for the two eras.  The support table defines valid source combinations. It is more specific than the rectangular shape of the physical array.

| Source era | Valid ages | Ages per county/sex/race | Race/ethnicity levels |
|---|---|---:|---|
| Before 2017 | `< 1`, single ages 1–84, `85 +` | 86 | asian, black, hispanic, other, white |
| 2017 onward | `< 1`, single ages 1–94, `95 +` | 96 | asian, black, hispanic, other, white |

Before 2017, Asian estimates were not separately reported; they were included in the source's "Other"
category. Asian nevertheless remains valid support and is represented by `NA` for those years. Do
not impute Asian counts or repartition the historical other counts during ingestion.

Age labels have interval meaning. For example, `86` denotes `[86, 87)` which is shorthand for beginnig at 86 and every
value up to but not inluding 87 and `85 +` denotes `[85, Inf)`. The union of labels across eras includes overlapping
intervals, so the global age semantics use a set with overlap information. Within a given year's supported rows, the
source age categories are non-overlapping. Preserve those categories: never manufacture ages 85–94 from an early `85 +`
count or replace later single ages with an ingested `85 +` aggregate.

## Three different kinds of cells

An observed count, including zero, is a source value. A supported `NA` is a valid combination whose
count is unavailable. A structural `NA` is an unsupported combination present only because an array
requires a rectangular grid. These are different cases even though the latter two are stored as `NA`.

For 2016, `85 +` is supported while `86` and `95 +` are structural. For 2017, `86` and `95 +` are
supported while `85 +` is structural. A structural missing value must not become zero. An observed
zero must not become missing.

`All` is a virtual aggregate, not a physical base-cube label. County data and a Texas state total also
must not be stored together under county-partition semantics: the state total overlaps every county.
The transformer can retain canonical `Texas` rows when explicitly requested, but this county-cube
workflow excludes them.

# The actual pipeline

```text
annual CSV files
  -> read_est_csv() for each file
  -> read_tdc_estimates_raw()
  -> transform_tdc_estimates()
  -> prepare_population_df()
  -> apply_completion_policy()
  -> validate_population_df()
  -> rectangularize_population_df()
  -> df_2_array()
  -> pa_write_poparray_cube()
  -> rebuild_poparray_registry()
  -> open_poparray()
```

`ingest_population()` orchestrates reading, transformation, preparation, completion, and validation.
It then calls `build_poparray_from_df()`, which rectangularizes the table, constructs the array, and
calls the HDF5 writer. The writer uses `HDF5Array::writeHDF5Array()` for the population dataset and
writes the metadata separately. Registry rebuilding is a subsequent step needed to discover a new
cube through `open_poparray()`.

The reader and transformer functions are user written and are found in `data-raw/tdc_estimates.r`. Age-label and support
helpers live in `data-raw/tdc_estimates_support.r`. G

# Prepare a reproducible session

Record package versions, source file names, checksums, acquisition dates, and any source revisions.
Keep the raw files immutable. 

```{r setup}
# Prevent automatic cube-store initialization while loading the development package.
Sys.setenv(TARR_POP_SKIP_CUBE_SETUP = "true")
library(tidyverse)
library(janitor)
library(data.table)
library(rage)
devtools::load_all()

# Capture the configured production root before selecting an isolated build root.
production_root <- tarr.pop::cube_path()
production_file <- file.path(production_root, "base", "tdc_estimates_county.h5")
production_file
sessionInfo()
```

`tarr::paths$population` is the local path convention used by the supplied script. It is not a
portable requirement: set `input_dir` explicitly for another computer. The local ingestion tooling
uses packages such as readr, janitor, forcats, and rage in addition to the package itself.

Find the source data files and place the file names in a list called "files".

```{r source-manifest}
input_dir <- file.path(tarr::paths$population, "Estimates", "Texas Demographic Center", "asre")
pattern <- "^20[1-2][0-9]_ASRE_Estimate_alldata\\.csv$"
files <- sort(list.files(input_dir, pattern = pattern, full.names = TRUE))
stopifnot(length(files) > 0L)
manifest <- data.frame(
  file = basename(files),
  year = as.integer(substr(basename(files), 1L, 4L)),
  md5 = unname(tools::md5sum(files))
)
stopifnot(!anyNA(manifest$year), !anyDuplicated(manifest$year))
manifest
```

The anchored pattern excludes zip archives and names that merely contain the expected suffix. If
several revisions exist for one year, choose the authoritative revision explicitly rather than
binding both. Update the file name policy when new years no longer match the current pattern.

## Load definitions without running the production script

Sourcing all of `tdc_estimates.r` executes its build with overwrite enabled. To inspect stages safely,
load its function definitions only. The following repository-only setup reuses the actual functions
instead of maintaining a second copy of their implementations in this guide.

```{r load-definitions}
tdc <- new.env(parent = globalenv())
sys.source("data-raw/tdc_estimates_support.r", envir = tdc)
wanted <- c("read_est_csv", "read_tdc_estimates_raw",
            "transform_tdc_estimates", "tdc_estimate_semantics")
for (expr in parse("data-raw/tdc_estimates.r")) {
  if (is.call(expr) && identical(expr[[1L]], as.name("<-")) &&
      is.symbol(expr[[2L]]) && as.character(expr[[2L]]) %in% wanted) {
    eval(expr, envir = tdc)
  }
}
stopifnot(all(vapply(wanted, function(nm) is.function(tdc[[nm]]), logical(1))))
```

Internal helpers referenced with `tarr.pop:::` below are useful for a maintainer's staged audit.
They are not a stable public API. A routine build should call the package exported function
`ingest_population()`.

# Read and audit the sources

`read_est_csv()` inspects the header before selecting column specifications. It handles `County`
versus `Area Name`, and `FIPS` versus `Area Code`. It cleans names, maps Anglo to white, removes the
non-Hispanic prefix and population suffix, and removes source total columns. It uses an existing year
column when present, otherwise the year in the file name is used. Therefore verify both agree rather than assuming
the file name always wins.

The current reader first reads the file fields as a character vector, then `read_tdc_estimates_raw()` removes a comma
and converts character columns to integer. A successful CSV parse alone cannot prove that this conversion succeeded.


# Transform and confirm geographic identity

```{r transform}
dims <- c("year", "area.name", "sex", "age.char", "race.eth")
counties <- setdiff(names(tarr.pop::county_fips), "Texas")
canonical <- tdc$transform_tdc_estimates(
  data.table::copy(raw), counties = counties, include_texas_total = FALSE
)
stopifnot(identical(names(canonical), c(dims, "population")),
          !anyDuplicated(canonical[, ..dims]),
          !any(canonical$population < 0, na.rm = TRUE),
          !"Texas" %in% as.character(canonical$area.name))
```

Pass a copy because the transformer changes a data.table by reference. It resolves source FIPS/Area
Code through the existing one-to-one county reference before discarding that identifier. Short
county codes are normalized to Texas county FIPS; code 123 resolves to `De Witt` even when the source
label is `DeWitt`. Unresolved geography identifiers cause an error before the whitelist can silently
drop them. State identifiers resolve to canonical `Texas`; exclusion is controlled explicitly by
`include_texas_total`, including when `counties = NULL`.

The wide counts are then melted, race/sex column names are split, age labels are interpreted through
rage, aggregates are removed, and remaining factor levels are retained in their intended order.
Do not reorder raw age factor levels alphabetically before this transformation. Check infancy and
upper-age labels directly after normalization.

# Build complete support and apply completion

```{r support-and-completion}
support <- tdc$tdc_estimate_support_table(years = manifest$year, counties = counties)
stopifnot(!anyDuplicated(support[, dims]), !anyNA(support[, dims]))
semantics <- tdc$tdc_estimate_semantics()
prepared <- tarr.pop:::prepare_population_df(
  data.table::copy(canonical), dims, drop_all = TRUE, data_col = "population"
)
completed <- tarr.pop:::apply_completion_policy(
  data.table::copy(prepared), dims, policy = "na", data_col = "population",
  support = data.table::as.data.table(support)
)
tarr.pop:::validate_population_df(completed, dims, allow_na = TRUE, data_col = "population")
completed[is.na(population), .N, by = .(year, race.eth)][order(year, race.eth)]
```

Support must list **all** valid source combinations for every input year, county, sex, supported age,
and race/ethnicity. A table containing only missing Asian rows is not complete support. The completion
function rejects observed dimensions outside explicit support and duplicate support dimensions.
With `"na"`, valid but absent cells receive `NA`; unsupported combinations do not enter this table.
The `"error"` policy rejects incomplete support, while `"zero"` fills missing valid cells with zero.
Use `"na"` for TDC; missing reporting does not establish a true population of zero.

The supplied semantics use partitions for year, county, sex, and race/ethnicity. Age uses interval
scale and set semantics with `85 +` recorded as overlapping. Setting `validated = TRUE` is a contract
claim, not evidence that the CSV was audited; it must be justified by the preceding checks.

# Rectangularize and construct the array

```{r rectangularize}
rectangular <- tarr.pop:::rectangularize_population_df(
  data.table::copy(completed), dims, data_col = "population"
)
arr <- tarr.pop::df_2_array(as.data.frame(rectangular), data_col = "population")
stopifnot(nrow(rectangular) == prod(dim(arr)),
          !anyDuplicated(rectangular[, ..dims]),
          identical(names(dimnames(arr)), dims))
```

Rectangularization takes the Cartesian product of dimension extents and left-joins the supported
observations. The physical array contains the union of age labels from all represented eras, so
structural `NA` cells are expected. `df_2_array()` requires exactly one row per physical coordinate;
it maps values to explicit indices instead of relying on row ordering or recycling. Factor levels
control the intended ordering.

`build_poparray_from_df()` performs these steps internally. They are shown separately here so the
first stage that changes a total can be located before writing a file.

# Verify every in-memory stage

```{r conservation}
table_stats <- function(x) {
  x <- data.table::as.data.table(x)
  list(total = sum(x$population, na.rm = TRUE),
       non_na = sum(!is.na(x$population)), na = sum(is.na(x$population)),
       by_year = x[, .(population = sum(population, na.rm = TRUE)), by = year][order(year)],
       by_county = x[, .(population = sum(population, na.rm = TRUE)),
                     by = .(area.name = as.character(area.name))][order(area.name)])
}
baseline <- table_stats(canonical)
for (stage in list(prepared, completed, rectangular)) {
  stats <- table_stats(stage)
  stopifnot(stats$total == baseline$total, stats$non_na == baseline$non_na,
            isTRUE(all.equal(stats$by_year, baseline$by_year, check.attributes = FALSE)),
            isTRUE(all.equal(stats$by_county, baseline$by_county, check.attributes = FALSE)))
}
stopifnot(baseline$total == source_total,
          identical(as.numeric(baseline$by_year$population),
                    as.numeric(source_year_totals$population)),
          sum(arr, na.rm = TRUE) == source_total,
          sum(!is.na(arr)) == baseline$non_na)

supported_na <- sum(is.na(completed$population))
structural_na <- nrow(rectangular) - nrow(completed)
stopifnot(sum(is.na(arr)) == supported_na + structural_na)
data.frame(stage = c("canonical", "completed", "rectangular", "array"),
           total = rep(source_total, 4),
           non_na = rep(baseline$non_na, 4),
           na = c(sum(is.na(canonical$population)), supported_na,
                  supported_na + structural_na, sum(is.na(arr))))
```

Canonical source NAs, if present, are already missing before completion. To count newly introduced
NAs separately, subtract existing canonical missing dimensions rather than labeling every completed
NA as new. Inspect missing combinations by year, county, age, and race. A whole missing county is a
geography or reader problem, not legitimate unreported population.

For the audited 2011–2024 files, the county snapshot checksum was **397,324,736**, with **2,999,232**
observed non-NA cells, **262,128** supported early Asian NAs, and **187,960** structural NAs. The
physical dimensions were 14 × 254 × 2 × 97 × 5. These are historical checks for that exact source
set, not permanent constants: revisions or added years require a new independently verified baseline.

# Write to an isolated cube store

Do not use `overwrite = TRUE` as the first step of an audit. The writer removes an existing file
before constructing its replacement; it is not a transactional swap.

```{r temporary-build}
build_root <- tempfile("tdc-cube-build-")
tarr.pop::set_cube_path(build_root, create = TRUE, persist = FALSE)
dir.create(file.path(build_root, "base"), showWarnings = FALSE)
temporary_file <- file.path(build_root, "base", "tdc_estimates_county.h5")
source_meta <- list(
  note = "Texas Demographic Center county estimates",
  population_type = "Estimate",
  source = "Texas Demographic Center, Estimates program"
)
tarr.pop::ingest_population(
  reader = function(...) data.table::copy(raw),
  transformer = tdc$transform_tdc_estimates,
  dims = dims, dim_semantics = semantics,
  filepath = temporary_file, series_id = "tdc_estimates_county",
  completion_policy = "na", drop_all = TRUE, source_meta = source_meta,
  time_dim = "year", area_dim = "area.name", support = support,
  overwrite = FALSE, data_col = "population",
  counties = counties, include_texas_total = FALSE
)
stopifnot(file.exists(temporary_file), file.info(temporary_file)$size > 0)
tarr.pop::rebuild_poparray_registry(build_root)
cube <- tarr.pop::open_poparray("tdc_estimates_county")
```

The reader closure reuses the already audited raw table. To demonstrate a fresh source read instead,
pass `reader = tdc$read_tdc_estimates_raw`, `input_dir = input_dir`, and `pattern = pattern`; those
arguments travel through `...` to the source functions. Both reader and transformer receive `...`,
so their signatures must accept the arguments used for that build.

The scratch root contains only the candidate cube. `persist = FALSE` changes the session option
without changing the user's saved path. Avoid calling `init_cubes()` on this scratch root merely to
make directories: initialization can copy bundled cubes and introduce duplicate series or an
existing target file. For a new permanent store, `init_cubes()` is the supported initialization tool;
inspect its resulting contents before selecting an output path.

# Inspect storage and verify the reopened cube

```{r verify-hdf5}
inventory <- rhdf5::h5ls(temporary_file)
paths <- paste0(inventory$group, "/", inventory$name)
stopifnot("/cube/population" %in% paths,
          all(c("/cube/metadata/roles", "/cube/metadata/source",
                "/cube/metadata/dimnames", "/cube/metadata/dim_semantics") %in% paths))
inventory
```

The population dataset is `cube/population`. Metadata includes dimension order and labels, role
names, data-column name, source information, series identifier, and fieldwise dimension semantics.
A successful write return does not prove that values and metadata can be recovered.

```{r verify-reopened}
verify_cube <- function(x, expected_array) {
  stopifnot(identical(dim(x), dim(expected_array)),
            identical(dimnames(x), dimnames(expected_array)),
            identical(tarr.pop::time_role(x), "year"),
            identical(tarr.pop::area_role(x), "area.name"),
            identical(tarr.pop::data_col(x), "population"),
            identical(lapply(tarr.pop::dim_semantics(x), S7::props),
                      lapply(semantics, S7::props)))
  src <- tarr.pop::get_source(x)
  stopifnot(all(vapply(names(source_meta), function(nm) {
    identical(src[[nm]], source_meta[[nm]])
  }, logical(1))), length(src$updated) == 1L, nzchar(src$updated))
  # EAGER diagnostic: read the entire small-to-moderate candidate cube once.
  # For larger cubes, compare chunks at matching dimensions instead.
  values <- as.array(x)
  stopifnot(identical(as.numeric(values), as.numeric(expected_array)))
  invisible(values)
}
reopened_values <- verify_cube(cube, arr)
```

The comparison includes every coordinate, zero, and NA, so it is stronger than comparing totals.
It also implies exact yearly and county conservation when the input array has passed the earlier
checks. For a large cube, replace the full realization with matching blocks or year/county slices;
compare all blocks, not merely a random sample. Opening, inspecting dimensions, and selecting labels
remain lazy; requesting values is where reads occur.

Representative checks should still be printed for review. For example, where both years are present:

```{r representative-dimensions}
if (all(c("2016", "2017") %in% dimnames(arr)$year)) {
  data.frame(
    coordinate = c("2016 / De Witt / female / 85+ / white",
                   "2016 / De Witt / female / 86 / white",
                   "2017 / De Witt / female / 85+ / white",
                   "2017 / De Witt / female / 86 / white",
                   "2017 / De Witt / female / 95+ / white",
                   "2016 / De Witt / female / <1 / asian"),
    value = c(arr["2016", "De Witt", "female", "85 +", "white"],
              arr["2016", "De Witt", "female", "86", "white"],
              arr["2017", "De Witt", "female", "85 +", "white"],
              arr["2017", "De Witt", "female", "86", "white"],
              arr["2017", "De Witt", "female", "95 +", "white"],
              arr["2016", "De Witt", "female", "< 1", "asian"])
  )
}
zero_row <- canonical[!is.na(population) & population == 0][1L]
if (nrow(zero_row) && !is.na(zero_row$population)) {
  coordinate <- lapply(zero_row[, ..dims], as.character)
  stopifnot(do.call(`[`, c(list(arr), coordinate, list(drop = TRUE))) == 0)
}
```

Also review De Witt totals and its 960 supported cells per later year, along with another county such
as Tarrant. For the audited 2020–2024 sources, De Witt totals were 19,824; 19,886; 19,660; 19,784;
and 20,242. Those observations must survive all stages, rather than reappearing as missing cells
introduced by completion.

## Print yearly and county conservation results

```{r final-conservation}
final_by_year <- data.frame(
  year = as.integer(dimnames(reopened_values)$year),
  population = vapply(dimnames(reopened_values)$year, function(year) {
    sum(reopened_values[year, , , , ], na.rm = TRUE)
  }, numeric(1))
)
yearly_comparison <- merge(as.data.frame(source_year_totals), final_by_year,
                           by = "year", suffixes = c(".source", ".cube"))
yearly_comparison$difference <- with(yearly_comparison, population.cube - population.source)
stopifnot(nrow(yearly_comparison) == nrow(manifest),
          all(yearly_comparison$difference == 0))
yearly_comparison

de_witt <- completed[as.character(area.name) == "De Witt",
  .(population = sum(population, na.rm = TRUE), cells = .N,
    supported_na = sum(is.na(population))), by = year][order(year)]
stopifnot(identical(as.numeric(de_witt$population), as.numeric(source_de_witt$population)),
          all(de_witt$cells == ifelse(de_witt$year < 2017L, 860L, 960L)))
de_witt
baseline$by_county[as.character(area.name) %in% c("De Witt", "Tarrant")]
```

For a different source or a deliberately restricted county selection, adapt these source-specific
checks to its declared support. Do not compare filtered observations with an unfiltered reference.

# Replace production only after verification

Print and confirm the actual configured production path; do not copy another computer's absolute
path from a report. Preserve a verified backup **outside** the registry-scanned `base` directory,
including its subdirectories. An HDF5 backup inside that tree can be discovered as a duplicate series.

The following is a deliberate publication step, not part of rendering this guide. It uses the normal
ingestion entry point again and then compares production with the verified candidate. It is safe to
proceed only after every previous check succeeds and no process is writing the same production file.

```{r production-rebuild}
production_file
backup_file <- NULL
if (file.exists(production_file)) {
  backup_dir <- file.path(dirname(production_root), "tdc_estimates_backups")
  dir.create(backup_dir, recursive = TRUE, showWarnings = FALSE)
  backup_file <- file.path(backup_dir, paste0(
    "tdc_estimates_county_", format(Sys.time(), "%Y%m%dT%H%M%S"), ".h5"
  ))
  stopifnot(!file.exists(backup_file),
            file.copy(production_file, backup_file, overwrite = FALSE),
            unname(tools::md5sum(production_file)) == unname(tools::md5sum(backup_file)))
}
tarr.pop::set_cube_path(production_root, create = TRUE, persist = FALSE)
tarr.pop::ingest_population(
  reader = function(...) data.table::copy(raw), transformer = tdc$transform_tdc_estimates,
  dims = dims, dim_semantics = semantics,
  filepath = production_file, series_id = "tdc_estimates_county",
  completion_policy = "na", drop_all = TRUE, source_meta = source_meta,
  time_dim = "year", area_dim = "area.name", support = support,
  overwrite = TRUE, data_col = "population",
  counties = counties, include_texas_total = FALSE
)
tarr.pop::rebuild_poparray_registry(production_root)
tarr.pop:::reset_poparray_cache()
production_cube <- tarr.pop::open_poparray("tdc_estimates_county")
production_values <- verify_cube(production_cube, arr)
stopifnot(identical(as.numeric(production_values), as.numeric(reopened_values)))
```

If the rebuild or verification fails, do not report success. Stop other use of the defective candidate,
retain diagnostics, and restore the verified backup if replacement damaged production. Rebuild the
registry and reset caches after a restoration as well. Close or discard previously opened objects
before replacing files they reference; do not assume a live lazy object tracks file replacement safely.
The metadata update date may change between builds, so compare provenance fields and valid update
records rather than requiring identical timestamps.

Restore the intended cube path when leaving a scratch-only session. Retain the source manifest,
yearly and county checksums, missing-cell classification, spot checks, session information, candidate
path, production path, and backup path as the build record.

# Troubleshooting and maintenance

| Symptom | Likely boundary to inspect | Action |
|---|---|---|
| County vanishes or becomes entirely missing | Source identifier mapping and whitelist | Inspect FIPS before filtering; never repair by zero fill |
| Counts become NA during reading | Character-to-integer conversion | Inspect original strings and conversion warnings |
| Observations are outside support | Source transformation or age-era support | Compare exact offending dimensions before changing support |
| Totals increase | Duplicate source revisions or aggregate rows | Remove overlap from the comparison and inspect duplicates |
| `df_2_array()` rejects row count | Missing rectangularization or incomplete table | Complete valid support, then rectangularize separately |
| Labels or values shift after conversion | Factor order or coordinate mapping | Compare explicit dimensions and dimnames |
| Unknown series when opening | Wrong active root or stale registry | Set the correct root and rebuild its registry |
| Duplicate series when opening | Multiple files with the same series ID | Move backups out of the scanned base tree; rebuild the registry |
| NA count seems excessive | Supported versus structural missingness | Classify against support, then inspect missing counties/years |
| Aggregation is refused | Overlapping labels or invalid semantics | Select a valid non-overlapping set; do not bypass guards |

For a new population source, keep the same division of responsibilities. Its reader handles source
schemas and parsing; its transformer resolves identities and constructs canonical observations; its
support helper states valid combinations; its semantics define safe operations. The generic pipeline
then validates, rectangularizes, writes, and reopens. A direct call to `df_2_array()` is suitable only
for a complete rectangular table already checked at that boundary.

Before releasing an ingestion change, run focused source and support tests, then `devtools::test()`.
Run `devtools::check(document = FALSE, vignettes = FALSE, error_on = "never")` to distinguish package
check issues from build accuracy. Existing unrelated example, documentation, or installed-test errors
must be reported separately; a full test-suite pass alone does not establish real-data conservation.
When adding fixtures, ensure installed-package tests do not unconditionally source files excluded by
`.Rbuildignore`.

Normal analysis should keep the verified `poparray` HDF5-backed and lazy, use named dimensions, and
retain roles and semantics through subsetting and aggregation. Full-cube realization belongs in an
explicitly budgeted build diagnostic, not in routine population analysis.
